---
title: "Using ggplot2 to visualize data"
author: "jdegen"
date: "May 17, 2018"
output: html_document
---

## What is ggplot2?

ggplot2 is a language created by Hadley Wickham, based on “The Grammar of Graphics” by Wilkinson et al (2005), for describing and creating plots (graphics). Plots can be built up iteratively and updated later.

[Here](http://www.rstudio.com/wp-content/uploads/2016/11/ggplot2-cheatsheet-2.1.pdf) is a useful "cheat sheet" overview of common ggplot functions. 

## The grammar of graphics

What is a graphic?
A mapping from **data** to **aesthetic attributes** (e.g., color, shape, size) of **geometric objects** (e.g., points, lines, bars).

#### A graphic 

- may contain statistical **transformations** of the data
- is drawn on a specific **coordinate system**
- can use **faceting** for generating the same plot for different subsets of the dataset

## Components of a graphic

#### Minimally:

- **data** and set of **aes**thetic mappings describing how to map variables in data to aesthetic attributes
- geometric objects (**geom**s) are what you see: points, lines, bars

#### Optional:

- statistical transformations (**stats**) summarise data (e.g., binning, counting observations)
- **scale**s map values in data space to values in aesthetic space
- **coord**inate system describes how data coordinates are mapped to graphic plane
- **facet**ing specification describes how to break data into subsets andreturn an individual plot for each subset (like conditioning)

```{r}
source("helpers.R")
library(languageR)
library(lme4)
library(tidyverse)
data(lexdec)
```

Set the background to be white instead of gray.
```{r}
theme_set(theme_bw())
```


It's always good to start out by understanding your response distribution. You *can* do this with a table. 
```{r}
table(lexdec$RT)
```

But some things are easier to understand by visualization. Here's a histogarm of the response distribution. 

```{r}
ggplot(lexdec, aes(x=RT)) +
  geom_histogram() 
```

Always add reasonable/informative axis labels.
```{r}

```

To save a plot, use ggsave().
```{r}
ggsave(file="graphs/rt_histogram.png",width=5,height=4)
```

Plot response time as a function of frequency.
```{r}

```



Plot RT as a function of frequency, by native language.
```{r}

```



Manually adjust colors with use of scales. All visual details (colors, shapes, line types, coordinate systems, etc) are controlled via scales.
```{r}

```

Plot mean untransformed RT by native language as barplot with error bars.
```{r}
lexdec$rawRT = exp(lexdec$RT)
agr = lexdec %>%
  group_by(NativeLanguage) %>%
  summarize(MeanRT = mean(rawRT), CI.Low = ci.low(rawRT), CI.High = ci.high(rawRT)) %>%
  mutate(YMin = MeanRT - CI.Low, YMax = MeanRT + CI.High)


```

Add individual data points to barplot to visualize participant variability.
```{r}

```
Add by-participant means to barplot to visualize variability.
```{r}

```

Plot log RT as a function of frequency by subject (individual variability in sensitivity to frequency). Useful: facets.   
```{r}

```

Scatterplot with an added line. Empirical observations against model predictions.
```{r}
m = lmer(RT ~ Frequency*NativeLanguage + FamilySize + (1+Frequency+FamilySize|Subject) + (1+NativeLanguage|Word),data=lexdec)
lexdec$PredictedRT = fitted(m)

ggplot(lexdec, aes(x=PredictedRT,y=RT)) +
  geom_point(size=.5) +
  xlab("Model predicted log response times") +
  ylab("Empirical log response times")
```